Two applications of logic programming to Coq

### 2020-11-02 papers/types2020/sec3.tex
Added label to section 3.
More about implementation

%\newpage
\section{Proof theory and proof certificates}

@@

\todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}
    \newcommand{\XXi}{{\color{blue}{\Xi}}}

  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash A : s}
{(A,s) \in \mathcal{A}
}
\qquad
\infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
{\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi;\Gamma\vdash k l : A}
      {\bc{\XXi}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma}
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we will never have occurrences of
applications of hypotetic assumptions. This means, at the term level, that
we don't need free variables in our grammar of terms.

\begin{align*}
t,u :=& k l \;|\; t t \;|\; \forall x : t.\, u \\
l :=& [] \;|\; (t l)
\end{align*}

The syntax will depend on the external set of inductive definitions available
in the context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where

Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
%
\todo{Explain relations with other term assignments. Note that we do not use explicit substitutions (we are cut free). Explain the choice of not having certificates represent proof terms. Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality): this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that arand inherited by Coq }

Figure~\ref{fig:kernel} contains the elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
%
% Notice that it is easy to show that no matter how the expert predicates
% are defined, if the goal \lsti{check Cert B} is provable in \lP then
% \lsti{B} must be a sound consequence of the program clauses stored in
% \dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

@@

\todo{What happened to the unfold certificate???}
% \todo{Show the dprolog tactic? have an example? could  be
% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}}
The notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

>>>>

%\newpage
\section{Proof theory and proof certificates}
\label{sec:three}

@@

\todo{Figure now uses Dale kernel-v2 --- needs further abstraction and explanation, see appendix}
\begin{footnotesize}
\begin{figure}
  \newcommand{\XXi}{{\color{blue}{\Xi}}}
  \newcommand{\bc}[5]{#1;#2\vdash #3 : #4 \Downarrow #5}
\[
% \infer{\XXi\vdash t = t}
% {%\eqExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash \true}
% {%\trueExpert{\XXi}
% }
% \qquad
% \infer{\XXi\vdash G_1\vee G_2}
% {\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
% }
% \qquad
% \infer{\XXi\vdash \exists x. G}
% {\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
% }
% \qquad 
\infer{\XXi;\Gamma\vdash A : s}
{(A,s) \in \mathcal{A}
}
% \qquad
% \infer{\XXi;\Gamma\vdash \lambda x:A.\,t : \forall x : A.\,  G}
% {\XXi;\Gamma, x : A\vdash t : G%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
% }
\qquad
\infer{\XXi_1;\Gamma\vdash k\ l : A}
      {\bc{\XXi_2}{\Gamma} l D A \quad k : D \in \Pscr_A\cup\Gamma\qquad \unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\bc {\XXi}{\Gamma} {[]} A A}{\trueExpert{\XXi}}{}\qquad
%  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{\XXi}{\Gamma}{(t\ ::\ l)} {\forall x:B.\, D} A}
  {\XXi_1;\Gamma\vdash t : A\qquad \bc{\XXi_2}{\Gamma}l {B[t/x]} A\qquad\prodExpert{\XXi}{\XXi_1}{\XXi_2}}{}
\]
%\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
\end{figure}
\begin{figure}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

@@

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we will never have occurrences of
applications of hypotetic assumptions. This means, at the term level, that we
don't need to have variables in our grammar of terms. Terms are always
applied to a (possibly empty) list of arguments, and as is usual in Pure Type
Systems we have the same grammar for terms and types, as shown in figure~\ref{fig:terms}.

\begin{figure}
\begin{align*}
t,u :=& k\ l \;|\; \forall x : t.\, u \\
l :=& [] \;|\; (t\ ::\ l)
\end{align*}
\caption{Grammar of terms}
\label{fig:terms}
\end{figure}

The syntax depends on the external set of inductive definitions available in
the context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where $D$ is of the form
$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
by $\mathcal{A}$ the set of arity declaration for the inductive types.

Figure~\ref{fig:augmented} contains a simple proof system \dots in which each
inference rule is augmented with an additional premise involving an
\emph{expert predicate}, a certificate $\Xi$, and possibly continuations of
certificates ($\Xi'$, $\Xi_1$, $\Xi_2$) with extracted information from
certificates.% (in the case of $\vee$ and $\exists$).
%
%
%
\todo{Explain relations with other term assignments. Note that we do not use
explicit substitutions (we are cut free). Explain the choice of not having
certificates represent proof terms. Explain that the unfold clause handles
defined atoms and defined connectives (conjunction, disjunction, existential
and also equality): this may be less efficient and proof-theoretically induce
pointless phase changes, but it is arguably very elegant e simplifies proof
terms that arand inherited by Coq }

Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate, and every rule
is directly mapped to a clause in the code. The argument of the clause is
either given by a \lsti{go} or a \lsti{bc} constructor, signaling whether the clause
is a goal-reduction or a backchaining one.

Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is the signature of the term constructors we use:
\begin{lstlisting}
  kind term   type.
  type prod   name -> term -> (term -> term) -> term.
  type app    list term -> term.
  type global gref -> term.
  type indt   inductive -> gref.
  type indc   constructor -> gref.
\end{lstlisting}

The constructor for products forms a term from a term and a \lP function from
terms to terms; when, in the implementation of the product-left rule, we
apply the second argument of \lsti{prod _ Ty1 Ty2} to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. The meaning of this
application is meta-level substitution, in the style of Higher Order Abstract
Syntax: one can think as if something of type \lsti{term -> term} were a
\lsti{term} with a hole in it for another \lsti{term}.

The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

%
% Notice that it is easy to show that no matter how the expert predicates
% are defined, if the goal \lsti{check Cert B} is provable in \lP then
% \lsti{B} must be a sound consequence of the program clauses stored in
% \dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

@@

% \todo{Show the dprolog tactic? have an example? could  be
% \lsti{Goal  exists R, insert 2 [0;1] R.} and we would move the code for insert here}
{\color{red}}
The notion of proof certificates
 is taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
In our case here, an FPC is a collection of \lP clauses that
provide the remaining details not supplied in Figure~\ref{fig:kernel}:
that is, the exact set of constructors for the \lsti{cert} type as
well as the exact specification of the expert predicates listed in
that figure.
%
Figure~\ref{fig:resources} displays two such FPCs,
both of which can be used to describe proofs where we bound
the number of occurrences of unfoldings in a proof.
%
For example, the first FPC
provides the experts for treating certificates that are constructed
using the \lsti{qheight} constructor.
%
As is easy to verify, the query \mbox{\lsti{(check (qheight 5) B)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC uses the constructor \lsti{qsize} (with two
integers) and can be used to bound the total number of instances of
unfoldings in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.
